Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2017-07-22 cabios-template/draft.tex
draft 2

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

@@

\begin{document}
\firstpage{1}

\subtitle{Subject Section}

\title[R packages for analyzing genome-wide data]{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author[Sample \textit{et~al}.]{Florian Privé\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France and \\
$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Motivation:} Genome-wide genetic studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge for the genetic community, severely slowing down genetic analyses. This increase in data size has led to the emergence of a range of specialized software for every part of the analysis pipeline. Yet, it is often difficult to choose which piece of software to use and how to combine all these software. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of the RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of our packages by analyzing a simulated Genome-Wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} https://privefl.github.io/bigstatsr/ and https://privefl.github.io/bigsnpr/\\
\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Consider first the basic tools necessary to perform a standard genome-wide analysis. Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}. Similarly, genome-wide analysis quality control software have been developed such as PLINK and Bioconductor GWASTools \cite[]{Gogarten2012}. Regarding computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, there are also several software available including different implementations of Principal Component Analysis (PCA) in EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}. Then, GWAS analysis itself depends on the genotype format, e.g. ProbABEL or SNPTEST for dosage data \cite[]{Aulchenko2010,Marchini2010}.  Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice \cite[]{Vilhjalmsson2015,Euesden2015}. As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions. Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.

Secondly, increasing size of genetic datasets is the source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers. This is particularly a burden for commonly used analysis languages such as R, Python and Perl. Solving the memory issues for these languages would give access to a broad range of tools for data analysis, already implemented by the scientific community. Hopefully, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read/write operations. Storing large matrices on disk and accessing them via memory-mapping is available in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.

@@

We are aware of another work that used memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008}. With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}. The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}. For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra \cite[]{Zeng2017,RSpectra2016}.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=160pt]{conversion+QC.pdf}}
\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
\end{figure}

Fast read/write operations from/to bed/bim/fam PLINK files are available. One should use PLINK to convert data to this format. In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure~\ref{fig:qc}). PLINK files are then read into a ``bigSNP'' object, which contains the genotype ``big.matrix'', a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``big.matrix''.

We developed a special ``big.matrix'' object, called ``BM.code'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage (use of one byte per element, 8 times less disk storage space than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''). With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{imputation.pdf}}
\caption{Imputation and reading functions available in package bigsnpr.}\label{fig:impute}
\end{figure}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{svd.pdf}}
\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array.}\label{fig:svd}
\end{figure}

For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions for the computation of a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition and the other on randomized projections (Figure~\ref{fig:svd}).

The function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}. FlashPCA2 and bigstatsr use the same PCA algorithm called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra. The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes the multiplication based on the ``big.matrix'' format, which enables parallel computations.

Fast algorithms for thinning SNPs  similar to algorithms provided in PLINK have been developed. For instance, thinning is mandatory when computing PCs of a genotype matrix \cite[]{Abdellaoui2013}. There are different options to thin SNPs based on Linkage Disequilibrium. The first option is known as pruning, which is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning.

@@

The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in these regions while computing PCs\footnote{https://goo.gl/8TngVE}. However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}. Finally, the algorithm keeps only consecutive outlier SNPs which is considered as the sign of long-range LD regions by the algorithm. Indeed, a long-range LD region would cause SNPs in this region to have strong weights (loadings) in the PCA and we can differentiate these from true outliers because they are consecutive. This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.

@@

For association purposes, statistical tests based on linear and logistic regressions are available. Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}.

$\forall j \in \{1, \dots, m\}$, 
\itemize{
\item for the linear regression: $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, where $m$ is the number of SNPs, $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender).
\item for the logistic regression: $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, where $\hat{p} = \mathbb{P}(Y = 1)$.
}\\
The hypothesis that is tested is $\beta^{(j)} = 0$ against the alternative $\beta^{(j)} \neq 0$.

The R packages also implement functions to compute Polygenic Risk Scores.

First, they implement the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in previous equations. Under the P+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value. The SNPs are first clumped (P) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Dudbridge2013,Chatterjee2013,Golan2014}.

@@

In this paper, we use two real datasets: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.

For computation timing purposes only, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same structure of Linkage Disequilibrium and Population Structure as the original dataset. To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also by replicating the Celiac dataset. \label{sec:rep}

@@

All the code used in this paper along with results, such as execution times and figures, are available as HTML notebooks in the Supplementary Data.

@@

We present the results for three different analyses. First, we illustrate the application of R packages bigstatsr and bigsnpr. Secondly, we compared the performance of the packages against standard software. Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes. We use three types of data: Celiac, a real case-control cohort, POPRES, a general population cohort and simulated datasets using real genotypes from the Celiac cohort. We compare the performance on two computers, a desktop computer with 64GB of RAM and 12 cores, and a laptop with only 8GB of RAM and 4 cores. For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

We performed an association study and computed a polygenic risk score for the Celiac cohort.

The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $1^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that was then imputed with our new method (see section~\ref{sec:impute}). We note that if we used a standard R matrix to store the genotypes, this data would take 32GB of memory. On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{celiac-pca}}
\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{celiac-gwas-cut}}
\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. The y-axis has been cut in order to not see only the very strong effects on chromosome 6.}\label{fig:gwas}
\end{figure}

@@

\begin{table}[!tpb]
\centering
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|cc|cc|}
\hline
Operation & Functions involved (without packages) & Execution time & Execution time & Functions involved (with packages) \\ 
\hline
Reading PLINK files & n/a & n/a & 5 / 20 sec & snp\_readBed \\
Pruning & plink -{}-exclude -{}-range -{}-indep-pairwise & 4 / 6 sec & 13 / 46 sec & snp\_indLRLDR \& snp\_pruning \\
\multirow{2}{*}{Computing 10 PCs} & plink -{}-extract -{}-make-bed & 2 sec & \multirow{2}{*}{45 / 136 sec} & \multirow{2}{*}{big\_randomSVD} \\
& flashpcaR::flashpca (FlashPCA2) & 6 / 7 min &  &  \\
GWAS (binary phenotype) & plink -{}-logistic -{}-covar & 5 min & 5 / 14.5 min & big\_univLogReg \\
\hline
Total & PLINK and flashpcaR & 11 / 12 min & 6 / 18 min & bigsnpr and bigstatsr \\
\hline
\end{tabular} 
\end{adjustbox}
\caption{Execution times with or without bigstatsr and bigsnpr for making a GWAS for the Celiac dataset. The first time is with the desktop computer and the second time is with the laptop computer (section~\ref{sec:overview}).} 
\label{tab:bench-gwas}
\end{table}

\begin{table}[!tpb]
\centering
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|cc|cc|}
\hline
Operation & Functions involved (without packages) & Execution time & Execution time & Functions involved (with packages) \\
\hline
GWAS (binary phenotype) & plink -{}-keep -{}-logistic -{}-covar & 4 / min & 3 / min & big\_univLogReg \\
Clumping & plink -{}-keep -{}-clump -{}-clump-p1
-{}-clump-p2 -{}-clump-r2 & 49 / sec & 9 / sec & snp\_clumping \\
PRS & plink -{}-keep -{}-extract -{}-score -{}-q-score-range & 9 / sec & 4 / sec & snp\_PRS \\
\hline
Total & PLINK & 5 / min & 3 / min & bigsnpr and bigstatsr \\
\hline
\end{tabular} 
\end{adjustbox}
\caption{Execution times with or without bigstatsr and bigsnpr for making a GWAS for the Celiac dataset. The first time is with the desktop computer and the second time is with the laptop computer (section~\ref{sec:overview}).} 
\label{tab:bench-prs}
\end{table}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{benchmark-pca}}
\caption{Benchmark comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA/EIGENSOFT) and package bigstatsr. It shows the computation time in minutes as a function of the number of samples. The computation corresponds to 10 Principal Components and 93,083 SNPs which remain after thinning.}\label{fig:bench-pca}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{precision-pca}}
\caption{Precision comparisons between randomized Partial Singular Value Decomposition available in FlashPCA2, FastPCA (fast mode of SmartPCA) and package bigstatsr. It shows the squared correlation between approximated PCs and ``true'' PCs (given by the slow mode of SmartPCA) of the celiac disease dataset (whose individuals have been repeated 1, 5 and 10 times).}\label{fig:prec-pca}
\end{figure}

@@

For the detection of long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the Celiac and POPRES datasets. For the POPRES dataset, the algorithm converged in two iterations. The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table~\ref{tab:lrldr-popres}). We compared the scores (PCs) obtained by this method with the ones obtained by removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}). For the Celiac dataset, we found 5 long-range LD regions and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping with removing of predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).

@@

For the fast imputation method based on XGBoost, we compared  the imputation accuracy and computation times on the POPRES dataset. The minor allele frequencies (MAFs) are approximately uniformly distributed between 0.05 and 0.5 (Figure~\ref{fig:maf}). We introduced missing values using a Beta-binomial distribution resulting in approximately 3\% of missing values. Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle (v21Jan17). Overall, our method made 4.7\% of imputation errors whereas Beagle made only 3.1\% but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less). For the Celiac dataset, our method took less than 6 hours to complete whereas Beagle didn't finish imputing chromosome 1 in 48 hours. We also show that our method's estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{error-impute}}
\caption{Number of imputation errors vs the estimated number of imputation errors by SNP. For each SNP with missing data, the number of imputation errors corresponds to the number of individuals for which imputation is incorrect. The estimated number of errors is a quantity that is returned when imputing with snp\_fastimpute, which is based on XGBoost \cite[]{Chen2016}.}\label{fig:error-impute}
\end{figure}

@@

We think that the two R packages and the corresponding data format could help researchers to develop new ideas and algorithms to analyze genome-wide data. For example, we wish to use these packages to train much more accurate predictive models than the standard P+T model currently in use. As a second example, multiple imputation has been shown to be a very promising method for increasing statistical power of a GWAS, and it could be implemented with the data format ``BM.code'', without having to write multiple files \cite[]{Palmer2016}.

\section*{Acknowledgements}

>>>>

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}

\linenumbers
\doublespacing
%\usepackage[authoryear]{natbib}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

@@

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\usepackage{babel}
\makeatother

\begin{document}

\title[R packages for analyzing genome-wide data]{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author[Privé \textit{et~al}.]{Florian Privé\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France and \\
$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.

$^\ast$To whom correspondence should be addressed.

\date{~ }

\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge for the genetic community, severely slowing down genetic analyses. Specialized software for every part of the analysis pipeline have been developed to handle large genomic data. However, combining all these software into a single data analysis pipeline might be technically difficult. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} https://privefl.github.io/bigstatsr/ and https://privefl.github.io/bigsnpr/\\
\textbf{Contact:} \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Consider first the basic tools necessary to perform a standard genome-wide analysis. Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK, available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}. Similarly, quality control software for genome-wide analysis have been developed such as PLINK and Bioconductor GWASTools \cite[]{Gogarten2012}. Computation of principal components (PCs) of genotypes is commonly performed to account for population stratification in association studies and there are several software available, including EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}. Then, implementation of GWAS analyses also depends on the genotype format, e.g. ProbABEL or SNPTEST for dosage data \cite[]{Aulchenko2010,Marchini2010}.  Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice, which provide prediction for quantitative traits or disease risks based on multiple variants \cite[]{Vilhjalmsson2015,Euesden2015}. As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions. Overall, this might be a brake on data exploration.
%Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.

Secondly, increasing size of genetic datasets is the source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers. This is particularly a burden for commonly used analysis languages such as R, Python and Perl. Solving the memory issues for these languages would give access to a broad range of tools for data analysis, already implemented by the scientific community. Hopefully, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping is available in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.

@@

We are aware the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008}. With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}. The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}. For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra \cite[]{RSpectra2016,Zeng2017}.

@@

Fast read/write operations from/to bed/bim/fam PLINK files are available so that one should convert data to this binary PLINK format. In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure~\ref{fig:qc}). PLINK files are then read into a ``bigSNP'' object, which contains the genotype ``big.matrix'', a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``big.matrix''.

We developed a special ``big.matrix'' object, called ``BM.code'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage. Indeed, each element is stored on one byte which is 8 times less disk storage space than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''. With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).

@@

@@

For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD, (Figure~\ref{fig:svd}).

The function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}. FlashPCA2 and bigstatsr use the same PCA algorithm called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra. The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the ``big.matrix'' format, which enables parallel computations.

Fast algorithms for thinning SNPs  similar to algorithms provided in PLINK have been developed. For instance, thinning is mandatory when computing PCs of a genotype matrix \cite[]{Abdellaoui2013}. There are at least 3 different options to thin SNPs based on Linkage Disequilibrium. The first option is known as pruning, which is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning.

@@

The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in long-range LD regions while computing PCs\footnote{https://goo.gl/8TngVE}. However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}. Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions by the algorithm. Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA. This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}. 
Statistical tests based on linear and logistic regressions are also available. For the linear regression, for each SNP $j$, a T-test is performed on $\beta^{(j)}$ where $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$ and $\hat{p} = \mathbb{P}(Y = 1)$.
\\
The hypothesis that is tested is $\beta^{(j)} = 0$ against the alternative $\beta^{(j)} \neq 0$.

The R packages also implement functions to compute Polygenic Risk Scores. 
First, they implement the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in previous equations. Under the P+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value. The SNPs are first clumped (P) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Dudbridge2013,Chatterjee2013,Golan2014}.

@@

In this paper, two real datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.

For comparing computation times of different software, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same structures of Population and Linkage Disequilibrium as the original dataset. To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also by replicating the Celiac dataset. \label{sec:rep}

@@

All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the Supplementary Data.

@@

We present the results for three different analyses. First, we illustrate the application of R packages bigstatsr and bigsnpr. Secondly, we compared the performance of the packages against standard software. Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes. We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. We compare the performance on two computers, a desktop computer with 64GB of RAM and 12 cores, and a laptop with only 8GB of RAM and 4 cores. For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

We performed an association study and computed a polygenic risk score for the Celiac cohort. 
The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $1^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that was then imputed with our new method (see section~\ref{sec:impute}). We note that if we used a standard R matrix to store the genotypes, this data would take 32GB of memory. On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB.

@@

@@

@@

@@

For the detection of long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the Celiac and POPRES datasets. For the POPRES dataset, the algorithm converged in two iterations. The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table~\ref{tab:lrldr-popres}). We compared the scores (PCs) obtained by this method with the ones obtained by removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}). For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping with removing of predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).

@@

For the fast imputation method based on XGBoost, we compared  the imputation accuracy and computation times on the POPRES dataset. The minor allele frequencies (MAFs) are approximately uniformly distributed between 0.05 and 0.5 (Figure~\ref{fig:maf}). We introduced missing values using a Beta-binomial distribution (Figure~\ref{fig:NA}) resulting in approximately 3\% of missing values. Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle (v21Jan17). Overall, our method made 4.7\% of imputation errors whereas Beagle made only 3.1\% but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less). For the Celiac dataset, our method took less than 6 hours to complete whereas Beagle didn't finish imputing chromosome 1 in 48 hours. We also show that our method's estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).

@@

We think that the two R packages and the corresponding data format could help researchers to develop new ideas and algorithms to analyze genome-wide data. For example, we wish to use these packages to train much more accurate predictive models than the standard P+T model currently in use when computing PRS. As a second example, multiple imputation has been shown to be a very promising method for increasing statistical power of a GWAS, and it could be implemented with the data format ``BM.code'', without having to write multiple files \cite[]{Palmer2016}.

\newpage
\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{conversion+QC.pdf}}
\caption{Conversion and Quality Control preprocessing functions available in package bigsnpr via system calls to PLINK.}\label{fig:qc}
\end{figure}
